---
title: |
  Climate Change and Aging: A systematic Review

# to produce blinded version set to 1
blinded: 0

authors: 
- name: Mathew E. Hauer
  # thanks: The authors gratefully acknowledge ...
  affiliation: Department of Sociology, Florida State University
  
- name: Kyle Rose
  affiliation: Department of Sociology, Florida State UniversityW

keywords:
- population aging; demography; climate impacts; mitigation; adaptation

bibliography: bibliography.bib
output: rticles::asa_article
---

```{r load, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
read_chunk('../R/SCRIPTS/000-libraries.R')
read_chunk('../R/SCRIPTS/101-DocumentWorkFlow.R')
read_chunk('../R/SCRIPTS/201-TextNumbers.R')
read_chunk('../R/SCRIPTS/102-HeatMapCrossTab.R')
read_chunk('../R/SCRIPTS/103-Map.R')

```

```{r libraries, include = FALSE}
```

```{r textnumbers, include=FALSE}
```

---
abstract:  'In this systematic review, we analyze the literature through Web of Science’s SCI-Expanded containing the words “aging” or “aged” and “climate change” and receiving at least four citations per year since publication (n = `r num_articles_terms` articles). After discarding irrelevant articles (ie, “aging infrastructure”), the `r num_articles_reviewed` remaining articles overwhelmingly (`r percent((num_temp_mort + num_temp_morb)/num_articles_reviewed)`) fall into two categories: temperature/mortality (n= `r paste(num_temp_mort, percent(num_temp_mort/num_articles_reviewed), sep = "; ")`) and temperature/morbidity (n = `r paste(num_temp_morb, percent(num_temp_morb/num_articles_reviewed), sep = "; ")`). However, many other important climate topics related to aging remain underdeveloped. Notably, adaptation (n = 50; 29%), vulnerability (n = 27; 16%), emissions/mitigation (n = 19; 11%), climate perceptions (n = 13; 8%), drought (n = 4; 2%), and food security (n = 2; 1%) remain understudied. Furthermore, more than half of the studies were conducted in the United States (n = `r paste(num_US, percent(num_US/num_articles_reviewed), sep = "; ")`), China (n = `r paste(num_china, percent(num_china / num_articles_reviewed), sep = "; ")`), Globally (n = `r paste(num_global, percent(num_global/num_articles_reviewed), sep = "; ")`), and Australia (n = 11; 6%), suggesting a paucity of information in the Global South (n = 11) where climate impacts will be greatest. There were more studies specifically on Spain (n = 12) than specifically on the entire African continent (n = 5). Finally, 27 articles (16%) offered projections in some form, most to the middle of the century. Gerontologists and aging scientists should look beyond the relationship between heat and mortality to offer a more holistic view of aging and climate change. Prospective analyses, as opposed to retrospective, could shed additional light on the link between aging and climate change.'
---

# Introduction

Two  seemingly immutable trends will crash head-on during this century: the global populace will continue to age and global climate change impacts will worsen as the century progresses. By the end of the century, when climate change impacts will be considerably more intense than today, the Global populace exposed to these impacts will be decidedly older, amplifying climate change impacts. Many of the anticipated impacts from climate change disproportionately impact the elderly versus younger, more vigorous age groups making these two trends particularly potent when taken together.

Consider  this: the Global populace will age universally. The global population aged 75+ is expected to grow from 271M people today to nearly 1.4B people by 2100 [@division_wpp2017]. Today, only Japan, Italy, and Germany have median ages in excess of 50. But by 2100, 54% of countries will be as old as or older than these countries are today. These demographic trends are well known [@lutz2008coming; @beard2016world].

While  the global, dramatic aging shift occurs, the world will continue to warm. Climate projections suggest global temperatures are likely increase more than 3 degrees Celsius by 2100 compared to the pre-industrial period [@arias2021climate]. These temperature increases will likely usher in increasingly frequent and destructive extreme weather events, more frequent droughts and wildfire risks, extreme heat, and substantially increase the burden on health services [@portner2022climate]. Low-lying coastal areas will experience increased coastal flooding and the submergence of many coastal cities appears likely [@kulp2019new]. Climate change will make the world of 2100 considerably more precarious compared to today, a precarity amplified by a globally aging populace.

Nearly every climate impact is heightened by age due to the older people's social vulnerability and physiological susceptibility. Older adults have increased social vulnerability to psychological stresses due to environmental change, reduced ability to adapt, limited transportation, reduced mobility, smaller social networks, lower incomes, chronic health problems, social isolation, cognitive decline, and general fragility [@kovats2008heat]. This elevated social vulnerability combines with older adult’s increased physiological vulnerability to extreme heat and cold, extreme weather events, and infectious diseases to create a biophysical cocktail of potential catastrophe. 

Additionally, climate change and aging intersect in multiple ways, beyond just climate impacts. Older adults play important roles in the mitigation or reduction of carbon emissions [@oneill_global_2010; @buchs_who_2013]. Adaptation to climate impacts greatly vary among older populations [@huang_is_2011;@guo_high_2012] and perceptions of climate risks can serve as barriers to effective adaptation [@hansen_perceptions_2011;@abrahamson_perceptions_2008]. Yet, as we show in this article, most research concerning aging and climate change tends to focus on the impact of extreme temperatures on older populations. A more holistic view of the aging and climate change literature could shed light on the varied relationships between older adults and global environmental change.

In this article, we conduct a systematic review of the literature surrounding climate change and aging. For a full submission, we will investigate commonalities along mitigation, adaptation, and retrospective vs. prospective studies.



# Methods and Materials

We use a systematic literature review to assess the literature on climate change and aging.

## Document selection
We used a keyword search on Clarviate’s Web of Science-expanded search engine using the Boolean operator “TS=(aging OR aged OR elderly) AND TS=(“climate change”).” We selected Web of Science due to its comprehensive scientific coverage of peer-reviewed literature. We conducted the search on September 7, 2022. This search retrieved an initial universe of `r txtInt(num_articles_all)` articles. We filtered these results to include articles of relatively high impact, defined subjectively as those articles with at least four citations per year (n = `r txtInt(num_articles_4cites)`). To further isolate those articles pertaining to aging and climate change, we further restricted our search to those articles containing the words (aging or aged or elderly) in the abstract (n = `r txtInt(num_articles_terms)`). 

We then reviewed these articles for relevance, discarding articles concerning “aging infrastructure” or “aging forests” to isolate articles on human aging and climate change. This yielded a total of `r txtInt(num_articles_reviewed)` articles included in this systematic review (**\autoref{fig-diagram}**).

```{r figureflowdiagram, echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("PRISMA Flow Diagram. \\label{fig-diagram}")}
```

## Document review
Following our document selection and screening, `r txtInt(num_articles_terms)` articles were retained for full review. We developed a questionnaire to survey these articles to document and characterize the primary topics of climate change. We developed this questionnaire to standardize the analysis, produce descriptive statistics, and examine trends. We coded all papers based on (1) the primary and (2) secondary climate effect studied, (3) the climate impact type (sensitivity, vulnerability, or exposure), (4) the climate impact studied (morbidity, mortality, etc.), if the article concerned (5) mitigation, (6) adaptation, (7) or perceptions, if the article included a projection (8), the historic time period (9), and the general area the study was conducted (10). Additionally, we gathered general information on authorship, publication year, and citation counts. We conducted an extensive full-text review of all (n = `r txtInt(num_articles_terms)`) articles using this questionnaire. We assessed the primary finding in articles where multiple climate impact types or climate impacts were studied. 

## Analysis

All (n = `r txtInt(num_articles_all)`) articles were retained for validation. All data were entered into an Excel spreadsheet. We used R to analyze the data and to produce descriptive statistics and visualizations.

# Data Availability
The underlying computer code and data that support the findings of this study are available in the Supplementary Material and have been deposited in Zenodo (DOI). 

# Results

<!-- ## Table: Key Points (Matt) -->

## Figure: Heat Map of Topics and Bar Chart (Kyle)
Figure 2 illustrates the co-occurrences of climate effects and impacts. The overwhelming majority of articles in our sample concerned temperature as a climate effect (n = `r paste(num_temp)`). Some examples of climate effects which were classified as temperature include heat (n = 32), heatwaves (n = 16), daily temperature range (n = 9) and ambient temperature (n = 6).

The large majority of articles in our sample concerned the climate impact on mortality (n =`r paste(num_mort)`) and morbidity (n = `r paste(num_morb)`). The morbidity category included articles which specifically observe "morbidity" (n = 18), hospital admissions (n = 9) and ambulance attendance (n = 1), among others. 

More than half of the articles in our sample which included climate effects were specifically about the impact of temperature on mortality (n = 67). The second most frequent climate effect and impact to be studied together was temperature and morbidity (n = 24). Our findings suggest that the most popular climate change features studied in relation to aging populations are temperature effects like heat index and temperature variability on morbidity and mortality metrics like hospital attendance or years of life lost.

There is also a small section of literature which concerns the effect of pollution on aging populations. (n = `r paste(num_pollution)`. Most of these, like temperature, concerned pollution's impact on either mortality (n = 10) or morbidity (n = 3).

There is a paucity of information on how droughts, extreme weather and sea-level rise will be impacting aging human populations. Cumulatively, `r paste(num_D_EW_SLR)` articles observed any of these effects of climate change.

Importantly, very few articles observe the relationships between these climate effects and human behaviors, such as migration, economic activity and climate behaviors & policies (n = `r paste(num_behave)`). This gap is particularly pressing because it speaks to aging populations' abilities to respond to climate change. It is conceivable that aging populations may respond to sea-level rise with migration, urban heat islands with outdoor green spaces, extreme weather events with improved household infrastructure, or seasonal temperature variation with voting behavior. All of these examples demonstrate how climate effects can impact human behavior and consequently adaptive capacity, and all remain understudied.
```{r heatmapcrosstab, echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("\\textbf{Heatmap of Common Characteristics.} Here we show the commonalities in the papers we reviewed. Most articles concerning Aging and Climate Change concern the relationship between temperature and mortality or temperature and morbidity. Very few studies touch on other climate impacts such as drought or Extreme weather. \\label{fig-heatmap}")}
```

<!-- ## Figure: Timeline of Papers and Citations (Matt) -->

<!-- ## Figure: Social Network (Kyle) -->

## Figure: Map of the Literature (Matt)
```{r mapandtimelines, echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Map of study location frequency. \\label{fig-map}")}
```

## Figure: Citation Network (Kye)
Figure X graphs the citation network of articles within our sample. The illustration suggests there is some engagement between articles, mostly those which concern the effect of temperature on aging populations (symbolized by circular nodes on the network). Among these articles engaging each other, most seem to be concerned with the climate impact on mortality, morbidity or no climate impact measure (symbolized by pink, purple and blue). A substantial portion of the articles in our sample are positioned at the edge of the citation network without ties to other articles. This suggests that a large number of articles in our sample are not citing others. Considering our sample consists of accounts cited 4 or more times per year, this preliminary finding suggests that many highly-cited articles about the impacts of climate change on aging populations do not engage other highly-cited articles about the same general topic. Next, this study tables the frequencies of articles concerning different climate effects and impacts rates of engagement with other articles within and between those groups (this needs edited).

Table X describes the frequencies which articles concerning different climate effects cite other articles concerning the same and different effects. 117 (87%) of the articles in our sample which concerned the effects of temperature cited other articles in our sample concerning the effects of temperature a total of 443 times. This finding suggests that articles concerning the effects of temperature on aging populations have a high rate of citing other highly-cited articles on the topic, suggesting a healthy level of communication within this community. This contrasts the rates which articles concerning pollution (33%), sea-level rise & flooding (0%), wildfires and droughts (17%), extreme weather events (0%) and none (21%) cite each other. The relatively low percentage of articles in these groups which cite other highly-cited articles within their group suggests low levels of communication. Based on this finding, this analysis suggests that prospective studies concerning these climate effects should engage and connect to the other highly-cited articles concerning the same climate effect.

Table X also allows us to assess the extent that these articles are citing other articles which observe different climate effects. Notably, 67% of articles concerning extreme weather events and 50% of articles concerning no specific climate effect cited articles which concerned different climate effects. Given the small count of articles in these groups, this finding may suggest that these communities are still emerging and are referencing the highly cited articles of different topics to establish their contributions. 

<!-- # Conclusion -->

<!-- # Discussion -->